---
title: "DESeq heatmap GSE90522"
author: "Juraj Michalik"
date: "2022-11-26"
output: html_document
---

------------------------------------------

LICENSE: MIT License for provided scripts.

All scripts are distributed to ease the building of the Atlas, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the MIT License for more details.

------------------------------------------

This script is used to load data from project from GEO NCBI GSE90522.
Requirements: DESeq2 and all R-packages below. R version was 4.2.1. Please be sure to install them before running this script.
R version 4.2.1.

Also, please go to this page and download count table: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE90522 (GSE90522_counts.tsv.gz file at the bottom of the page).
Unpack it. Name should be 'GSE90522_counts.tsv'. The file should be in same directory as this script.

```{r, include=FALSE}
library(DESeq2)
library(pheatmap)
library(dplyr)
library(gplots)

library(biomaRt)
library(AnnotationHub)
```

Functions for annotations.

```{r}
# adds annotation to object
# Get current Mus Musculus database from AnnotationHub if not downloaded already
getEnsMm102 <- function(){
  aHu <- AnnotationHub()
  aHuQuery <- query(aHu, pattern = c("Mus Musculus", "EnsDb", '102'))
  EnsMm102 <- aHuQuery[[1]]
  return(EnsMm102)
}

entreid2symbols <- function(entrezids){
  EnsMm102 <- getEnsMm102()
  symbol.vector <- mapIds(EnsMm102,
                          keys = entrezids,
                          column = "SYMBOL",
                          keytype = "ENTREZID",
                          multiVals = "first")
  # rename only if symbol is a thing
  symbols.vector <- ifelse(is.na(symbol.vector), entrezids, symbol.vector)
  return(symbols.vector)    
}
```

Make meta data.

```{r}
cols.data <- data.frame(Sample_ID = paste('Sample', seq(1, 11, 1), sep = '.'),
                        Description = c('Naïve CD8 T cells sample 1',
                                        'Virtual memory CD8 T cells sample 1',
                                        'Naïve CD8 T cells sample 2',
                                        'Virtual memory CD8 T cells sample 2',
                                        'Naïve CD8 T cells sample 3',
                                        'Virtual memory CD8 T cells sample 3',
                                        'Naïve CD8 T cells sample 4',
                                        'Virtual memory CD8 T cells sample 4',
                                        'True memory CD8 T cells sample 1',
                                        'True memory CD8 T cells sample 2',
                                        'True memory CD8 T cells sample 3'),
                        Cell.type = c('Naïve', 'Virtual memory', 'Naïve', 'Virtual memory', 'Naïve', 'Virtual memory', 'Naïve', 'Virtual memory',
                                      'True memory', 'True memory', 'True memory'))
```

Count matrix.

```{r}
count.matrix <- read.csv('GSE90522_counts.tsv', sep = '\t', row.names = 1)
```

Row data. This includes translation from ENTREZ to Gene symbol.

```{r}
row.data <- count.matrix[,1,drop=F]
count.matrix <- count.matrix[,-1]
```

Biomart to add Symbols to row.data.

```{r}
row.data$symbols <- entreid2symbols(rownames(row.data))
```

Check if demanded genes are in there.

```{r}
c('Klrk1', 'Ccr5', 'Gcnt1', 'Itga4', 'Il12rb1', 'Il18r1', 'Il18rap') %in% row.data$symbols
```

Now create DESeq object, with cell type as variable of interest in design.

```{r}
dds <- DESeqDataSetFromMatrix(countData = count.matrix,
                              colData = cols.data,
                              rowData = row.data,
                              design= ~ Cell.type)

rownames(dds) <- row.data$symbols
```

This is not needed, just test.

```{r}
dds <- DESeq(dds)
```

Perform vst (variance stabilizing transformation).

```{r}
vst <- vst(dds, blind=FALSE)
```

Draw desired heatmap.

```{r}
png('GSE90522_Heatmap.png', width = 720, height = 480)

subassay.vst <- assay(vst)[c('Klrk1', 'Il12rb1', 'Il18r1', 'Il18rap', 'Gcnt1', 'Itga4', 'Ccr5'),]
colfunc <- colorRampPalette(c("blue", "gray90", "red"))

pheatmap(subassay.vst, border = 'NA', scale = 'row', color=colfunc(64), cluster_rows = F,
         annotation_col = as.data.frame(colData(dds)['Cell.type']),
         main='Heatmap of selected genes for \n True Memory, Virtual Memory and Naïve cells', cex = 1.15 )

dev.off()
```